---
title: 'Tutorial for R package TreatmentSelection '
output:
  html_document:
    theme: united
---

#  {.tabset .tabset-fade .tabset-pills}

## Install the package 

This tutorial uses the Treatment Selection package to analyze the example data provided in the package.

First, you need to download and install the package:

```{r, eval=FALSE, message = FALSE, warning = FALSE}
install.packages("TreatmentSelection")
```

If you would like the current version of the package directly from github, use the `devtools` package:

```{r, eval=FALSE}
if (!require("devtools")) install.packages("devtools")
devtools::install_github("mdbrown/TreatmentSelection")
```

First, load the example data set for this package called `tsdata`. Four markers are included in the data example, a ''weak'' and a ''strong'' marker ($Y1$ and $Y2$ respectively), along with a weak/strong discrete markers. 

```{r, warning = FALSE}
library(TreatmentSelection)
set.seed(12321)

data(tsdata)

tsdata[1:5, ]
```


## Create TrtSel objects 


Once we have the package and our data loaded into R, we need to create a treatment selection R object using the function `trtsel`. This function takes as inputs a data.frame of treatment indicators, adverse event status, marker values, and other optional information. Once we have created this object, we can then use it to plot risk/treatment effect curves, estimate summary measures, and check model calibration. 

`tsdata` contains indicators for event and treatment. There are also two continous biomarkers `Y1` and `Y2` along with dichotomous versions of the two markers. First let's create a `trtsel` object using the marker `Y1`, and take a look at it's contents:


```{r}

trtsel.Y1 <- trtsel(event ~ (Y1)*trt, 
                    treatment.name = "trt", 
                    data = tsdata, 
                    study.design = "randomized cohort",
                    link = "logit", 
                    default.trt = "trt all")

trtsel.Y1
```


As we see above, the object contains information about the study design, model fit, fitted risks given treatment, and estimated treatment effect for each individual. 

Now create a `trtsel` object using a discrete marker. 

```{r}
# Y2_disc = as.numeric(Y2>0)

trtsel.Y2_disc <- trtsel(event ~ trt*(Y2_disc), 
                         treatment.name = "trt",
                         data = tsdata, 
                         study.design = "randomized cohort", 
                         link = "logit")
```


See `?trtsel` for more information. Now that we have created trtsel objects, we can plot, evaluate, calibrate and compare them. 

## Plot  

### Use the plot function  


Plot risk curves:

```{r}
plot(trtsel.Y1, 
            main = "Y1: Oncotype-DX-like marker", 
            plot.type = "risk", 
            ci = "horizontal", 
            conf.bands = TRUE, 
            bootstraps = 50,       #more bootstraps should be run than this in practice!
            trt.names = c("chemo.", "no chemo."), 
            show.marker.axis = FALSE)
```


For a binary marker, we calculate vertical confidence bands:


```{r}
tmp <- plot(trtsel.Y2_disc,
                   main = "Discrete version of Y2", 
                   plot.type = "risk", 
                   ci = "vertical", 
                   conf.bands = TRUE, 
                   offset = 0.01, 
                   bootstraps = 50, 
                   trt.names = c("chemo.", "no chemo."))
```


`tmp` is now a list with elements `plot` that holds the ggplot output, and `ci.bounds` which holds the information regarding the confidence bounds. 


```{r}
tmp$ci.bounds
```

#### Treatment effect curves

We can also plot the distribution of treatment effects. 


```{r}
plot(trtsel.Y1, 
            plot.type = "treatment effect", 
            ci = "horizontal", 
            conf.bands = TRUE, 
            bootstraps = 50)
```

```{r}
plot.trtsel(trtsel.Y2_disc, 
            plot.type = "treatment effect", 
            conf.bands = TRUE, 
            bootstraps = 50)
```

#### Selection impact plot

Plot 'selection impact' curves, which show the estimated event rate if different proportions of observations where treated based off the marker of interest.

```{r}
plot.trtsel(trtsel.Y1, 
            plot.type = "selection impact", 
            ci = "vertical", 
            conf.bands = TRUE, 
            bootstraps = 50)
```


## Evaluate marker performance  

Calculate summary measures of marker performance along with bootstrap confidence intervals.

```{r}
tmp <- evaluate(trtsel.Y1, bootstraps = 50)
tmp
```

```{r}
# access the estimates
tmp$estimates
```

```{r}
# discrete marker
evaluate(trtsel.Y2_disc, bootstraps = 50)
```

## Assess model calibration 

Currently, model calibration is only available for continuous markers. 

```{r}
calibrate(trtsel.Y1, 
                 groups = 10, 
                 plot = "treatment effect", 
                 trt.names = c("chemo.", "no chemo."))


```

See `?calibrate.trtsel` for more plot options. 


## Compare markers 

To compare markers, the trt and event labels must be identical for the two markers. Plots can not be generated if comparing a discrete marker with a continuous marker. 


```{r}
# trtsel object for the stronger marker 2
trtsel.Y2 <- trtsel(event~trt*(Y2 + Y1), 
                    treatment.name = "trt", 
                    data = tsdata, 
                    default.trt = "trt all")

# Compare the markers based on summary measures
mycompare <- compare(trtsel1 = trtsel.Y1, 
                     trtsel2 = trtsel.Y2, 
                     model.names = c("Model 1", "Model 2"), 
                     bootstraps = 50, 
                     plot = TRUE, 
                     ci = "vertical", 
                     offset = 0.01, 
                     conf.bands = TRUE)
mycompare
```


```{r}
## Compare two discrete markers Y1_disc = as.numeric(Y1>mean(Y1))
trtsel.Y1_disc <- trtsel(event~Y1_disc*trt, 
                         treatment.name = "trt", 
                         data = tsdata, 
                         study.design = "randomized cohort", 
                         link = "logit")


compare.trtsel(trtsel1 = trtsel.Y1_disc, 
               trtsel2 = trtsel.Y2_disc, 
               ci = "vertical", 
               offset = 0.2, 
               bootstraps = 50, 
               plot = TRUE, 
               conf.bands = TRUE, 
               annotate.plot = FALSE)
```

See `?compare.trtsel` for more options.

## Additional features 

### Including fitted risks 

Alternative to including a marker and fitting a logistic model, the user can specify fitted risks for trt = 0 and trt = 1. In this case, no model fitting will be implemented and all bootstrap confidence intervals will be conditional on the provided fitted model. 

```{r}
#calculate model fit
mymod <- glm(event~trt*(Y2), data= tsdata, family = binomial("logit"))

tsdata$fitted.t0 <- predict(mymod, newdata=data.frame(trt = 0,Y1 = tsdata$Y1, Y2 = tsdata$Y2), type = "response")
tsdata$fitted.t1 <- predict(mymod, newdata=data.frame(trt = 1,Y1 = tsdata$Y1, Y2 = tsdata$Y2), type = "response")

myfitted.trtsel <- trtsel( event~trt, treatment.name = "trt", 
                         data = tsdata,
                         fittedrisk.t0 = "fitted.t0",
                         fittedrisk.t1 = "fitted.t1",
                         study.design = "randomized cohort", 
                         default.trt = "trt all")

```

We can now use this `trtsel` object just as before, but confidence intervals will be smaller because we do not account for the variation due to model fitting. 

```{r}
plot.trtsel(myfitted.trtsel, bootstraps = 50, plot.type = "risk",
            ci = "horizontal", show.marker.axis = FALSE)
```


### Simple function to estimate summary measures


```{r}


mod <- glm(event~trt*Y1_disc,  data = tsdata, family = binomial())

tsdata.0 <- tsdata; 
tsdata.0$trt = 0 
tsdata.1 <- tsdata;
tsdata.1$trt = 1
delta.hat <- predict(mod, newdata= tsdata.0, type = "response") - predict(mod, newdata= tsdata.1, type = "response")

trtsel_measures(event = tsdata$event, trt = tsdata$trt, trt.rule = 1- tsdata$Y1_disc, trt.effect = delta.hat )

```

## References

Janes H, Brown MD, Huang Y, Pepe MS. 2014. An Approach to Evaluating and Comparing Biomarkers for Patient Treatment Selection. **International Journal of Biostatistics**